A Kochen-Specker system has at least 22 vectors

### 2014-12-14 paper/main.tex
fix another parpic
remove lines for algorithmic (they were broken). fix some other things

% begin vertical rule patch for algorithmicx
% http://tex.stackexchange.com/questions/144840/
\makeatletter
\newcommand*{\algrule}[1][\algorithmicindent]{\makebox[#1][l]{\hspace*{.5em}\vrule height .75\baselineskip depth .25\baselineskip}}%

\newcount\ALG@printindent@tempcnta
\def\ALG@printindent{%
    \ifnum \theALG@nested>0%
        \ifx\ALG@text\ALG@x@notext%
            % do nothing
            \addvspace{-3pt}%
        \else
            \unskip
            % 
            \ALG@printindent@tempcnta=1
            \loop
                \algrule[\csname ALG@ind@\the%
                    \ALG@printindent@tempcnta\endcsname]%
                \advance \ALG@printindent@tempcnta 1
            \ifnum \ALG@printindent@tempcnta<\numexpr\theALG@nested+1\relax%
            \repeat
        \fi
    \fi
    }%
% 
\patchcmd{\ALG@doentity}{\noindent\hskip\ALG@tlm}%
        {\ALG@printindent}{}{\errmessage{failed to patch}}
\makeatother
% end vertical line thingy

@@

In~\cite{aow11}, Arends, Wampler and Ouaknine discuss several
computer-aided methods
to test embeddability of a graph.  None of these methods could decide
for all graphs considered, whether they were embeddable or not.
% TODO give a small description of their deficiencies
% TODO give intermediate computation.

\piccaption{One of the two minimal non-embeddable graphs \label{fig:unemb-10-2}}
\parpic[r]{%
\begin{tikzpicture}[thick,scale=1.4,%
        p/.style={circle, draw, fill=white,%
                        inner sep=0pt, minimum width=5pt}]
            \draw
                \foreach \x in {90,210,330} {
                    (\x:0.7) -- (0:0)
                    (\x-30:1.3) -- (\x:0.7)
                    (\x+30:1.3) -- (\x:0.7)
                    (\x-30:1.3) -- (\x+30:1.3) 
                    (\x+30:1.3) -- (\x-30+120:1.3) 
                }
                (90:0.7) node[p]{}
                (210:0.7) node[p]{}
                (330:0.7) node[p]{}
                (0:0) node[p]{}
                (90+30:1.3) node[p]{}
                (90-30:1.3) node[p]{}
                (210+30:1.3) node[p]{}
                (210-30:1.3) node[p]{}
                (330+30:1.3) node[p]{}
                (330-30:1.3) node[p]{}
                (90:0.7) node[above]{$w$}
                (210:0.7) node[above]{$v$}
                (330:0.7) node[above]{$x$}
                (0:0) node[below]{$z$}
                (90+30:1.3) node[above]{$p_3$}
                (90-30:1.3) node[above]{$p_4$}
                (210+30:1.3) node[below]{$p_1$}
                (210-30:1.3) node[left]{$p_2$}
                (330+30:1.3) node[right]{$p_5$}
                (330-30:1.3) node[below]{$a$} ;
\end{tikzpicture}}

We propose a new method,
which for all graphs we considered,
could decide
whether they were embeddable or not.
First we give a pen-and-paper example.
\begin{proposition}\label{prop:unemb-10-2}
The graph in Figure~\ref{fig:unemb-10-2}
is not embeddable.
\end{proposition}
\begin{myproof}
Suppose it is embeddable.
Consider~$p_1$.
It is orthogonal to both~$a$ and~$v$.
Since $a$ and~$v$ are not collinear,
$p_1$ must be collinear to~$v \times a$,
the cross-product of~$v$ and~$a$.
Similarly, $p_2$ is collinear to~$v \times p_1 = v \times (v \times a)$.
Continuing in this fashion,
we see that
\begin{equation}\label{eq:ue1}
    a \text{ is collinear to }
    x \times (x \times( w \times (w\times (v \times (v \times a))))).
\end{equation}
Now, we may assume that~$z=(0,0,1)$ and $x=(1,0,0)$.
Thus: $v=(v_1,v_2,0)$;
$w = (w_1,w_2,0)$
and $a = (0, a_2,a_3)$ for some~$-1 \leq v_1,v_2,w_1,w_2,a_2,a_3 \leq 1$,
with~$v_1^2+v_2^2 = 1$; $w_1^2+w_2^2=1$ and~$a_2^2 + a_3^2=1$.
Now, \eqref{eq:ue1} becomes:
\begin{equation*}
\begin{pmatrix}
0\\
a_2\\
a_3\\
\end{pmatrix}
\text{ is collinear to }
\begin{pmatrix}
0\\
-a_2 v_1w_2 (v_1w_1 + v_2w_2) \\
-a_3 (v_1^2 w_1^2 + v_1^2 w_2^2 + v_2^2w_1^2 + v_2^2w_2^2)\\
\end{pmatrix}.
\end{equation*}
Consequently
\begin{align*}
    v_1w_2 (v_1w_1 + v_2w_2)
& = v_1^2 w_1^2 + v_1^2 w_2^2 + v_2^2w_1^2 + v_2^2w_2^2 \\
& = (v_1^2 + v_2^2)w_1^2 + (v_1^2 + v_2^2)w_2^2 \\
& = w_1^2 + w_2^2 \\
&= 1.
\end{align*}
Since~$v$ and~$w$ are not collinear,
we have by Cauchy-Schwarz $|\left<v,w\right>| < 1$. Thus
\begin{equation*}
1 > |v_1w_2\left<v,w\right>| = |v_1w_2(v_1w_1+v_2w_2)| = 1,
\end{equation*}
which is a contradiction.  Apparently the graph is not embeddable.
\end{myproof}

@@

\begin{algorithmic}[5]
    \While{there are unassigned vertices}
        \State pick an unassigned vertex~$v$
        \State assign~$V(v)=v$ 
        \State mark~$v$ as free
        \While{there are unassigned vertices adjacent to two
                different assigned vertices}
            \State pick such a vertex~$w$ adjacent to the
                    assigned ~$w_1$ and~$w_2$
            \State assign~$V(w)=V(w_1) \times V(w_2)$
            \State mark edges~$(v,w_1)$ and~$(v,w_2)$ as accounted for
        \EndWhile
    \EndWhile
    \pagebreak[2]
    \For{each pair of vertices~$(v_1, v_2)$}
        \If{$(v_1,v_2)$ is not an edge}
            \State record requirement:~``$V(v_1)$
                    is not collinear to $V(v_2)$''
        \EndIf
    \EndFor
    \For{each edge~$(v_1,v_2)$ not accounted for}
        \State record requirement:~``$V(v_1)$ is orthogonal to~$V(v_2)$''
    \EndFor
\end{algorithmic}

@@

With this method, we have decided in a day the embeddability
of every squarefree graph with minimal vertex order three of
less than 15, except for one.\footnote{
A list of all squarefree graphs with minimal vertex order three of
less than 15 vertices together with their embeddability can
be found here:
\url{http://kochen-specker.info/smallGraphs/}.
The graph for which we could not determine
embeddability can be found
here: \url{http://kochen-specker.info/smallGraphs/4d4b3f4b3f603f47414641654953625f3f.html}.
}
In particular:
\begin{comp}\label{comp:unemb20}
    Every squarefree graph of minimal vertex order three
    that is not 010-colorable
    of order less than or equal to 20
    contains, as a subgraph, one of the following three graphs:
    \begin{center}
    \begin{tikzpicture}[thick,scale=1.0,
        p/.style={circle, draw, fill=white,
                        inner sep=0pt, minimum width=5pt}]
            \draw
                \foreach \x in {90,210,330} {
                    (\x:0.7) node[p]{} -- (0:0) node[p]{}
                    (\x-30:1.3) node[p]{} -- (\x:0.7)
                    (\x+30:1.3) node[p]{} -- (\x:0.7)
                    (\x-30:1.3) -- (\x+30:1.3) 
                    (\x+30:1.3) -- (\x-30+120:1.3) 
                }
            ;
        \end{tikzpicture}
        \qquad
    \begin{tikzpicture}[thick,scale=1.0,
        p/.style={circle, draw, fill=white,
                        inner sep=0pt, minimum width=5pt}]
            \draw
                \foreach \x in {0,180} {
                    (\x+15:1.3) node[p]{} -- (\x+90:0.9) node[p]{}
                    (\x+15:1.3) -- (\x+45:0.5) node[p]{}
                    (\x+45:0.5) -- (\x+90:0.9)

@@

            ;
        \end{tikzpicture}
    \end{center}
    These three graphs are unembeddable.  The left and middle graph
    are the only minimal unembeddable squarefree graph.
For the first graph, we have proven directly that it is unembeddable.
See Proposition~\ref{prop:unemb-10-2}.
For the second graph, we also have a similar direct proof. The third graph is shown to not be embeddable using our algorithm.

Every squarefree graph of minimal vertex order three that is
not~$010$-colorable of order 21 contains an unembeddable subgraph.%
\footnote{A list of these graphs together with their unembeddable
subgraphs, can be found here:
\url{http://kochen-specker.info/candidates/}.
The source code for this computation can be found
at~\texttt{code/comp2.py} of~\cite{GH}.
}
\end{comp}

>>>>

@@

In~\cite{aow11}, Arends, Wampler and Ouaknine discuss several
computer-aided methods
to test embeddability of a graph.  None of these methods could decide
for all graphs considered, whether they were embeddable or not.
% TODO give a small description of their deficiencies
% TODO give intermediate computation.
We propose a new method,
which for all graphs we considered,
could decide
whether they were embeddable or not.
First we give a pen-and-paper example.
\piccaption{One of the two minimal non-embeddable graphs \label{fig:unemb-10-2}}
\parpic[r]{%
\begin{tikzpicture}[thick,scale=1.4,%
        p/.style={circle, draw, fill=white,%
                        inner sep=0pt, minimum width=5pt}]
            \draw
                \foreach \x in {90,210,330} {
                    (\x:0.7) -- (0:0)
                    (\x-30:1.3) -- (\x:0.7)
                    (\x+30:1.3) -- (\x:0.7)
                    (\x-30:1.3) -- (\x+30:1.3) 
                    (\x+30:1.3) -- (\x-30+120:1.3) 
                }
                (90:0.7) node[p]{}
                (210:0.7) node[p]{}
                (330:0.7) node[p]{}
                (0:0) node[p]{}
                (90+30:1.3) node[p]{}
                (90-30:1.3) node[p]{}
                (210+30:1.3) node[p]{}
                (210-30:1.3) node[p]{}
                (330+30:1.3) node[p]{}
                (330-30:1.3) node[p]{}
                (90:0.7) node[above]{$w$}
                (210:0.7) node[above]{$v$}
                (330:0.7) node[above]{$x$}
                (0:0) node[below]{$z$}
                (90+30:1.3) node[above]{$p_3$}
                (90-30:1.3) node[above]{$p_4$}
                (210+30:1.3) node[below]{$p_1$}
                (210-30:1.3) node[left]{$p_2$}
                (330+30:1.3) node[right]{$p_5$}
                (330-30:1.3) node[below]{$a$} ;
\end{tikzpicture}}
\par\noindent\parbox{\textwidth}{%
\begin{proposition}\label{prop:unemb-10-2}
The graph in Figure~\ref{fig:unemb-10-2}
is not embeddable.
\end{proposition}}
\begin{myproof}
Suppose it is embeddable.
Consider (the point associated to)~$p_1$.
It is orthogonal to both~$a$ and~$v$.
Since $a$ and~$v$ are not collinear,
$p_1$ must be collinear to~$v \times a$,
the cross-product of~$v$ and~$a$.
Similarly, $p_2$ is collinear to~$v \times p_1 = v \times (v \times a)$.
Continuing in this fashion, walking the circumference of the graph,
we find
\par\vskip-\lastskip % HACK to prevent eq. to mess with parpic
\begin{equation}\label{eq:ue1}
    a \text{ is collinear to }
    x \times (x \times( w \times (w\times (v \times (v \times a))))).
\end{equation}
Now, we may assume that~$z=(0,0,1)$ and $x=(1,0,0)$.
Thus: $v=(v_1,v_2,0)$;
$w = (w_1,w_2,0)$
and $a = (0, a_2,a_3)$ for some~$-1 \leq v_1,v_2,w_1,w_2,a_2,a_3 \leq 1$,
with~$v_1^2+v_2^2 = 1$; $w_1^2+w_2^2=1$ and~$a_2^2 + a_3^2=1$.
Now, \eqref{eq:ue1} becomes:
\begin{equation*}
\begin{pmatrix}
0\\
a_2\\
a_3\\
\end{pmatrix}
\text{ is collinear to }
\begin{pmatrix}
0\\
-a_2 v_1w_2 (v_1w_1 + v_2w_2) \\
-a_3 (v_1^2 w_1^2 + v_1^2 w_2^2 + v_2^2w_1^2 + v_2^2w_2^2)\\
\end{pmatrix}.
\end{equation*}
Consequently
\begin{align*}
    v_1w_2 (v_1w_1 + v_2w_2)
& = v_1^2 w_1^2 + v_1^2 w_2^2 + v_2^2w_1^2 + v_2^2w_2^2 \\
& = (v_1^2 + v_2^2)w_1^2 + (v_1^2 + v_2^2)w_2^2 \\
& = w_1^2 + w_2^2 \\
&= 1.
\end{align*}
Since~$v$ and~$w$ are not collinear,
we have by Cauchy-Schwarz $|\left<v,w\right>| < 1$. Thus
\begin{equation*}
1 > |v_1w_2\left<v,w\right>| = |v_1w_2(v_1w_1+v_2w_2)| = 1,
\end{equation*}
which is a contradiction.  Apparently the graph is not embeddable.
\end{myproof}

@@

\begin{algorithmic}[5]
    \While{there are unassigned vertices}
        \State pick an unassigned vertex~$v$
        \State assign~$V(v)=v$ 
        \State mark~$v$ as free
        \While{there are unassigned vertices adjacent to two
                distinct assigned}
            \State pick such a vertex~$w$ adjacent to the
                    assigned ~$w_1$ and~$w_2$
            \State assign~$V(w)=V(w_1) \times V(w_2)$
            \State mark edges~$(v,w_1)$ and~$(v,w_2)$ as accounted for
        \EndWhile
    \EndWhile
    \pagebreak[2]
    \For{each pair of vertices~$(v_1, v_2)$}
        \If{$(v_1,v_2)$ is not an edge}
            \State record requirement:~``$V(v_1)$
                    is not collinear to $V(v_2)$''
        \EndIf
    \EndFor
    \For{each edge~$(v_1,v_2)$ not accounted for}
        \State record requirement:~``$V(v_1)$ is orthogonal to~$V(v_2)$''
    \EndFor
\end{algorithmic}

@@

With this method, we have decided in a day the embeddability
of every squarefree graph with minimal vertex order three of
less than 15, except for one.\footnote{
A list of all squarefree graphs with minimal vertex order three of
less than 15 vertices together with their embeddability can
be found here:
\url{http://kochen-specker.info/smallGraphs/}.
The graph for which we could not determine
embeddability can be found
here: \url{http://kochen-specker.info/smallGraphs/4d4b3f4b3f603f47414641654953625f3f.html}.
}
In particular:
\begin{comp}\label{comp:unemb20}
Every squarefree graph of minimal vertex order three that is
not~$010$-colorable of order less than or equal 21
contains an unembeddable subgraph.%
\footnote{A list of these graphs together with their unembeddable
subgraphs, can be found here:
\url{http://kochen-specker.info/candidates/}.
The source code for this computation can be found
at~\texttt{code/comp2.py} of~\cite{GH}.}
In particular, those of order less than or equal 20
    contains, as a subgraph, one of the following three graphs:
    \begin{center}
    \begin{tikzpicture}[thick,scale=1.0,
        p/.style={circle, draw, fill=white,
                        inner sep=0pt, minimum width=5pt}]
            \draw
                \foreach \x in {90,210,330} {
                    (\x:0.7) node[p]{} -- (0:0) node[p]{}
                    (\x-30:1.3) node[p]{} -- (\x:0.7)
                    (\x+30:1.3) node[p]{} -- (\x:0.7)
                    (\x-30:1.3) -- (\x+30:1.3) 
                    (\x+30:1.3) -- (\x-30+120:1.3) 
                }
            ;
        \end{tikzpicture}
        \qquad
    \begin{tikzpicture}[thick,scale=1.0,
        p/.style={circle, draw, fill=white,
                        inner sep=0pt, minimum width=5pt}]
            \draw
                \foreach \x in {0,180} {
                    (\x+15:1.3) node[p]{} -- (\x+90:0.9) node[p]{}
                    (\x+15:1.3) -- (\x+45:0.5) node[p]{}
                    (\x+45:0.5) -- (\x+90:0.9)

@@

            ;
        \end{tikzpicture}
    \end{center}
    These three graphs are unembeddable.  The left and middle graph
    are the only minimal unembeddable squarefree graph.
For the first graph, we have proven directly that it is unembeddable.
See Proposition~\ref{prop:unemb-10-2}.
For the second graph, we also have a similar direct proof. The third graph is shown to not be embeddable using our algorithm.
\end{comp}
